Searching for standards of fairness in the transportation justice literature

### 2023-11-15 just-transportation/just-transportation.qmd
First polished version: PRE-RELEASE VERSION

---
title: "Searching for standards of fairness in the transportation justice literature: A scoping review"
subtitle: "Searching for standards of fairness"
author:
  - name:
      given: Anastasia
      family: Soukhov
    affiliations:
      - name: McMaster University
        department: School of Earth, Environment and Society
    email: soukhoa@mcmaster.ca
  - name:
      given: Ignacio
      family: Tiznado-Aitken
    affiliations:
      - name: University of Toronto Scarborough
        department: Department of Geography and Planning
    email: i.tiznadoaitken@utoronto.ca
  - name:
      given: Matthew
      family: Palm
    affiliations:
      - name: University of North Carolina Chapel Hill
        department: Department of City and Regional Planning
    email: palmmatt@unc.edu
  - name:
      given: Steven
      family: Farber
    affiliations:
      - name: University of Toronto Scarborough
        department: Department of Geography and Planning
    email: steven.farber@utoronto.ca
  - name:
      given: Antonio
      family: Paez
    affiliations:
      - name: McMaster University
        department: School of Earth, Environment and Society
    email: paezha@mcmaster.ca
bibliography: All_papers_in_extracted.bib
format: docx
---

@@

```{r, eval=FALSE}
Pflow_chart %>%
  DiagrammeRsvg::export_svg() %>% 
  charToRaw() %>% rsvg::rsvg_png(file="figures/Pflow_chart.png")
```

@@

                            df_conc <- data.frame(
                              id=rep(Consensus_only_analysis$`Covidence #`, lengths(df_conc)),
                              conc=unlist(df_conc)) %>% 
                              mutate(conc = ifelse(conc == "Inequitable exposure to externalities", "Ineq. externalities", conc),
                                     conc = ifelse(conc == "Sufficientarism / Capabilities approach", "Emerg. theories", conc),
                                     conc = ifelse(conc == "Transport-related exclusion", "TR exclusion", conc))

@@

NUM_standards_conc <- standards_conc %>% group_by(id) %>% summarize(count = n())
NUM_ENTRIES_standards_conc <- NUM_standards_conc$count %>% sum()

@@

```{r, message=FALSE}
standards_conc <- standards_conc %>% 
  group_by(conc, std_mcat) %>% summarize(total = n())

standards_conc <- standards_conc %>%
  mutate(total_perc_in_conc = total/sum(total),
         total_perc_in_all = total/NUM_ENTRIES_standards_conc)
sums_concs <- standards_conc %>% group_by(conc) %>% summarize(
  total_perc_in_all = sum(total_perc_in_all),
  total_perc_in_conc = 1) %>% mutate(std_mcat = "TOTALS")

@@

sums_std_mcat <- standards_conc %>% group_by(std_mcat) %>% summarize(total_perc_in_all = sum(total_perc_in_all),
  total_perc_in_conc = 1) %>% mutate(conc = "TOTALS")

@@

totals_conc_table_flex <- 
  # compose(totals_conc_table_flex,
  #         i = 1, j = c(1,2), part = "header", value = as_paragraph(c("Conceptualizations", "Conceptualizations"))) %>%
  prepend_chunks(flextable(totals_standards_conc_table %>% select(-c("Standards"))),
                 part = "body", i = c(1,2,3,4,5), j = 1, as_b(c("Vertical equity: ", "TR exclusion: ", "Wellbeing: ", "Spatial equity: ", "Horizontal equity: "))) %>% 
  prepend_chunks(part = "body", i = c(1,2,3,4), j = 2, as_b(c("Ineq. externalities: ", "Rights: ", "Emerg. theories: ", "Utilitarian: "))) %>% 
  # merge_h(part="header", i=1) %>% #header merge
  merge_at(i=5, j=1:2) %>%
  # align( align="center", part="header") %>%
  delete_part("header") %>%
  border_outer() %>%
  border_inner_h() %>%
  autofit()

totals_standards_table_flex <- 
  # compose(totals_standards_table_flex,
  #         i = 1, j = c(1), part = "header", value = as_paragraph(c("Standards"))) %>% 
  prepend_chunks(flextable(totals_standards_conc_table %>% select("Standards")),
                 part = "body", i = c(1,2,3,4), j = 1, as_b(c("Population: ", "Opportunity: ", "Infrastructure: ", "Environmental +: "))) %>%
  merge_at(i= 4:5, j= 1) %>%
  delete_part("header")%>%
  border_outer() %>%
  border_inner_h() %>%
  autofit()
```

```{r}
standards_conc_percentages_plot <- ggplot(totals_standards_conc, 
       aes(x=std_mcat, y=conc, fill= total_perc_in_all 
           #width = rep(10, each=50)
           ))+ 
  geom_tile() + 
  scale_fill_gradient2(low = "#AFAFED", mid= "#3F3FD4", high = "#191970", midpoint = 0.2, limits = c(0,.4), na.value = "#FCB5AC") +
  scale_x_discrete(limits=(totals_standards_conc$std_mcat)[order(totals_standards_conc$group)][1:5]) + 
  scale_y_discrete(limits=(totals_standards_conc$conc)[order(totals_standards_conc$group)] ) +
  geom_text(aes(label = total_perc_in_all %>% percent(accuracy = 0.1)), color = "white", size = 2.5) +
  geom_vline(xintercept=4.5, linewidth=1.5, color = "#191970") +
  geom_hline(yintercept = 43.5, linewidth=1.5, color="#191970")+
  theme_classic() +
  coord_flip()+
  theme(legend.position = "none",
        axis.ticks = element_blank(),
        axis.text.x=element_text(angle=20,hjust=1, colour = "black"),
        axis.text = element_text(size=7, colour = "black"),
        axis.title = element_blank())
#+ ggtext::geom_richtext(aes(x=5.385, y=23.5, label="Conceptualizations"), fill="grey", size=2) +
#  ggtext::geom_richtext(aes(x=3, y=49, label="Standards"), angle = 270, fill="grey", size=2) 
# standards_conc_percentages_plot
# ggsave("figures/standards_conc_percentages_plot.png")
```

@@

We need a super nice cover.

@@

Modern societies provide people with varying degrees of mobility and accessibility, from low wage Amazon workers’ depending on infrequent bus service, to Jeff Bezos taking private flights into space. Societies implicitly tolerate different levels of inequality in access, though these standards are occasionally made legible through policy standards. For example, U.S. transit agencies are required to set and adhere to maximum levels of disparity in service quality between minority and non-minority routes as defined by the U.S. Civil Rights Act. In contrast to prior reviews on how to measure inequality in transportation systems, this report is concerned with the implied or explicit standards that are used to judge whether the measured inequalities are fundamentally “fair” or unacceptable. The present study achieves this objective by scanning the state of academic knowledge in defining “fairness” in the transport domain.

@@

**Methods**
The authors conduct a scoping review of the literature. The review involves entering combinations of equity-related keywords into academic databases to identify relevant studies. The team initially identified 6,832 references from three diverse databases that they shrunk to a final sample using a four-step process:

1.	Removing 611 duplicate studies.
2.	Having a team of trained students review the remaining titles and abstracts for relevance, with 4,511 studies excluded for not sufficiently addressing transportation equity. 
3.	Reviewing the full-texts of the remaining 1,711 studies. At least two trained students reviewed each document, with authors breaking any ties. This resulted in 1,545 additional studies being excluded. At this stage, studies could be excluded if they did not include any standard for, or conceptualization of, equity.
4.	Extracting data on the "When", "Where", "Who", "What", and "How" of transport equity in each study. This included noting methods used, standards suggested, and inferring conceptualizations of equity that may connect to broader notions of justice from the remaining 165 studies.

**Results**
**When and Where: the contexts of justice**. Most (60%) studies deploying standards for equity are published on the Global North, including cases from Europe, the U.S., Canada, Australia, New Zealand, Japan, and Israel.  These studies focus on groups seen as disadvantaged within national contexts, with standards related to national policies or issues. In contrast, research on the Global South tends to adopt international standards in analyses, such as the U.N.’s Sustainable Development Goals (SDGs). Within the Global South, studies most commonly cover Asia, followed by Latin America. Studies in the latter region focus disproportionately on mobility barriers and the financial affordability of transportation. Studies pertaining to Africa are numbered, though they cover a broader scope of issues including informal transportation, citizens’ accessibility needs, and infrastructure development needs. 
Across the globe, the reviewed studies apply equity standards overwhelmingly to urban contexts (85%) and are concerned with a focused set of issues around. Studies on rural contexts are broader in the topics examined, from the equity of ferry service in the Philippines to the impacts of road construction on remote Amazonian communities.

@@

<!--We suggest: Who and What: the tools of mobility. I like our more-->
**What: the tools of mobility.** Most studies deploying equity standards apply them to either transit (37%) or pedestrian modes (25%). Many transit analyses are multimodal in nature, comparing the access transit provides against the access automobiles provided as an implicit standard for comparison. Others combine transit with non-auto modes to examine access for households without vehicles. These studies focus on delivering equitable mobility, sometimes in context of also meeting climate goals. However, they often fail to include explicit minimums of service that should be provided to different populations. In contrast, rights-based, infrastructure-focused studies apply principles of universal design in the study of transport systems and related infrastructure.

<!--we suggest: What: the benefits of mobility and accessibility. I like ours more; there is a distinction. -->
**What: the benefits of mobility.** Studies measure either movement (i.e., trips taken, trip quality, etc.) or the potential for movement (i.e. accessibility to specific destinations, or generally), with most studies focused on the latter. Over a quarter of the accessibility focused papers do not focus on access to specific destinations, but instead on the accessibility of the trip (i.e., level of service, barriers of infrastructure) or bundles of destinations. Studies that center particular destinations that should be reached/accessed include: employment, healthcare, grocery stores, and greenspace. Access to community support or places of worship, as well as childcare activities are less common in the reviewed literature.

<!--we suggest: "How": concepts of justice and standards of fairness. I like ours more - there is a distinction. -->
**How: concepts and standards of justice.** The most common standards are defined using access to opportunities cut offs (37.2%), opportunity standards, or thresholds related to the perceptions or activities of specific populations (35.8%), population standards. The latter often includes standards informed by public health, such as the number of physical activities undertaken in a week and perceived life satisfaction. Opportunity standards are evaluated through horizontal equity and spatial equity lenses, while population measures are evaluated through well-being frameworks. Another category that emerged in the literature are infrastructure-based standards, which include adherence to universal design guidelines or specific benchmarks for road network characteristics; they are less common in the literature comparatively and are often framed in rights-based terms, particularly as they relate to achieving universal design. But infrastructure standards around sustainable infrastructure (i.e. block lengths or distance between transit stops) have also been evaluated through vertical, horizontal and spatial lenses. The distribution of studies with respect to the kinds of standards deployed, and the broader concepts of justice used, are presented in the Figure below:

@@

From what does and does not exist in the state of knowledge today, the authors challenge practitioners, researchers, advocates, and government to commit to five calls to action:

@@

The standards of equity used in research are often weakly linked to any legally, intellectually, or publicly accepted conceptualizations of justice. Many standards are arbitrary. For example, the 15-minute city sets a threshold that is easy to remember, but is this really what is needed to achieve sufficient sustainable mobility for urban residents, as opposed to a 10 or 20 minute city?  And what conceptualization of human need should determine what services are essential enough to need to be within this threshold, and for whom?

@@

Mixed-methods research can help government set meaningful standards by better linking quantitative thresholds with qualitative assessments of well-being, freedom of movement and perceived access, among other dimensions. Yet most of the reviewed research is purely quantitative or, to a much lesser degree, purely qualitative. The constant deployment of metrics Lorenz Curves of Gini coefficients without research on where such measures need to be to improve quality of life (i.e., a standard), will not help governments with limited resources achieve the best possible equity outcomes within their respective contexts. Mobilizing Justice’s Priority Populations working group was inaugurated to advance new research in this area.

<!-- We suggested:
Call 3: Data availability as a matter of justice-->
Call 3: Data availability as a matter of justice.

Several areas of critical importance, such as access to care and community, are under-researched, and the authors suspect this relates to lack of data on these kinds of destinations. Furthermore, data issues prevent the application of many methods of equity analyses in the Global South. The production, availability, transparency, and ownership of data is a matter of justice on which governments and NGOs should act. In the Canadian context, Mobilizing Justice’s National Survey working group will advance data practices by conducting the first national survey on transportation poverty.

@@

Linking standards to lived experiences and individual outcomes can help increase their likelihood of adoption by governments. Further, they can help decision makers understand the role of mobility in improving policy outcomes in other domains like nutrition, physical and mental health, education, and human development.  These links can help make the standards used in research and policy more comprehensible and relatable to community members and people with lived experience navigating unequal mobility systems. Mobilizing Justice launched it’s Data Driven Standards working group to advance research in this area.

@@

Approximately 10% of the reviewed studies assessed interventions. Delivering new equity-focused policies and programs that do not actually advance equity harms communities, metastasizes mistrust, and delegitimates future efforts to promote transportation justice. Furthermore, policies that benefit one group of people in one context may not benefit other groups in other situations. A deep breadth of equity evaluations is needed to avoid failure and guide policymakers towards interventions that will work for the communities they serve. Mobilizing Justice launched its Innovative Pilots working group to begin filling this gap, through further work is needed.

@@

If you're interested in the motivation behind this report, read the first section, [Justice in transportation?](#sect1)

@@

- **Fairness**: is achieved when a collective and individuals are treated equally and without bias as defined by their morals and values.
- **Equity**: the distribution of benefits and burdens of things among the population, with a particular emphasis on those with the least advantage and quality in their outcomes.
- **Justice**: mutually reinforced by the tools of fairness and equity, justice is a multi-dimensional political ideal attained when people "give and receive whatever they are due" [@jaggarPhilosophicalChallengesGlobal2009, pp. 1-2]. It ceases to exist when there are persons or groups that are denied "access to the opportunities they need to lead a meaningful and dignified life" [@karnerTransportationEquityTransportation2020, p. 440]. The process towards justice is fluid and can be identified through questions that answer what is the context, subject and object of justice, and what is the rationale for justice.
- **The temporal-spatial context of justice**: the answers to "Where?" and "When?". What regions, within what countries, spatial contexts, point of time, and processes of planning are impacting transport systems?
- **The subjects of justice**: the answer to "Who?" is the focus of justice.
- **The objects of justice**: the answer to "What?" is the focus of justice; what are the things that characterise (in)justice. 
- **The methods for justice determination**: the answer to "How?" is justice defined and markers towards it tracked.
- **The rationale to justice**: the "Why?". The rationale connects with the context, subject, object and methods used for justice determination.
- **Equity standards**: are a tool for answering this distributive question: how do we allocate burdens and benefits and to whom? Standards are something that when operationalized effectively define what is fair.
- **Accessibility**:  
- **Capabilities approach**:  
- **Equality**:  
- **Mobility**:  
- **Policy**:  
- **Social exclusion**:  
- **Space-time convergence**:  
- **Sufficientarianism**:  
- **Theories of justice**:  
- **Utilitarianism**:  
Etc.,
<!--NOTE: not too sure how many of these terms we need.. or if we need a glossary at all. Perhaps we can ask peer-reviewers about this need.-->

@@

# Justice in transportation? {#sect1}

@@

The fiery wake of a rocket could be seen ascending, piercing the morning sky above West Texas. It was July 20, 2021, and on board was a small group of four passengers that included Jeff Bezos, the founder of Amazon and then the world's richest person [@harwoodBezos2021]. The mission that day was among the first-ever private suborbital passenger flights, and the adventure (described as "intense" by one of the passengers) lasted a total of 10 minutes and 10 seconds [@harwoodBezos2021]. In addition to intense, the undertaking was expensive: a seat for the flight had previously been auctioned for no less than $28 million USD [@griffinPerson2021]. Before even 10:00 am (EDT), when Bezos was back from his excursion, he took time to declare that this was the "[b]est day ever". To reporters covering the event he said "I want to thank every Amazon employee and every Amazon customer, because you guys paid for all of this" [@johnsonRoadToSpace2021].

Meanwhile, firmly grounded on planet Earth, the employees that Bezos thanked for his suborbital escapade were struggling with some very mundane problems of their own, and none as lofty as conflicting schedules that prevented them from flying in rockets [@griffinPerson2021]. According to reports, people employed directly or indirectly by Amazon for warehouse or delivery work had, for years, been treated to "inhumane" conditions [@fungEuropean2018; @scottAmazon2019; @greeneAlgorithm2021], and subjected to surveillance on the job, degrading schedules that drove drivers to urinate in bottles, crushing demands for productivity quotas that led to injury, all the while facing little or no job security. From this perspective, they were treated less with gratitude and more as disposable inputs to feed Amazon's earnings and consumers' demands [@tungAmazon2020; @BBCNews2021; @reeseTeamsters2022; @middletonRobot2023].

>>>>

---
title: "Searching for standards of fairness in the transportation justice literature"
subtitle: "Searching for standards of fairness"
format:
  mj-report-pdf: 
    keep-tex: true
doctype: "Report (Draft)"

# Author information. You can add or delete authors here
author:
  - name:
      given: Anastasia
      family: Soukhov
    affiliations: 
      - name: McMaster University
        department: School of Earth, Environment and Society
    email: soukhoa@mcmaster.ca
  - name:
      given: Ignacio
      family: Tiznado-Aitken
    affiliations: 
      - name: University of Toronto
        department: Department of Geography and Planning
    email: i.tiznadoaitken@utoronto.ca  
  - name:
      given: Matthew
      family: Palm
    affiliations: 
      - name: The University of North Carolina at Chapel Hill
        department: Department of City and Regional Planning
    email: palmmatt@unc.edu  
  - name:
      given: Steven
      family: Farber
    affiliations: 
      - name: University of Toronto
        department: Department of Geography and Planning
    email: steven.farber@utoronto.ca  
  - name:
      given: Antonio
      family: Páez
    affiliations: 
      - name: McMaster University
        department: School of Earth, Environment and Society
    email: paezha@mcmaster.ca

# Date: default is last modified but could be a string.  
date: last-modified

# Write the acknowledgments; remember to use a tab at the beginning of each paragraph
acknowledgements: | 
  This report would not have been possible without the help and inspiration of many individuals not listed as authors. In the early stages of this report we received invaluable help from specialists at the University of Toronto as we learned about various approaches to literature reviews, particularly the Research Services and Liaison Librarians. These experts provided guidance in cornering down the topic, identifying guidance material on literature synthesis and relevant databases, and eased us into the world of knowledge synthesis. Though our path diverged from a traditional scoping review, the protocol published on September 2022 (found [here](https://osf.io/rsb92)) was followed with only minor changes, with only a reduction in the databases searched due to resource constraints. We also wish to acknowledge the many hours of work by undergraduate and graduate research assistants who poured through title and abstract screening to help determine full-text eligibility at the earliest stages of our search. The volume of literature initially reviewed was greater than any of us had experienced before, but important to be as inclusive as possible. For this support, we are deeply thankful to Jan Domalaon, Dancel Gayle, Eryn Maloney, Mahdis Moghadasi, Anika Munir, and Subaita Refaaf.
  The views and opinions expressed here, as well as any equivocations and/or awkwardness are the authors' and the authors' alone.

# Place the cover image and the toc image in folder `images`
cover_image: cover-picture.png
cover_image_caption: "Source: Marco Verch (https://www.flickr.com/photos/160866001@N07/50597690137/)"
toc_image: toc-picture.png
toc_image_caption: "Source: Progressive Charlestown (https://www.progressive-charlestown.com/2020/01/amazon-drivers-die-for-you.html)"
# Information for generating citation
citation:
  type: report
  doi: "10.somenumber/document.00010"
  issued: last-modified
  url: https://example.com/summarizing-output
  number: "MJ-0001"
bibliography: bibliography.bib
---

@@

```{r, eval=FALSE}
Pflow_chart |>
  DiagrammeRsvg::export_svg() |> 
  charToRaw() |> rsvg::rsvg_png(file="figures/Pflow_chart.png")
```

@@

                            df_conc <- data.frame(
                              id=rep(Consensus_only_analysis$`Covidence #`, lengths(df_conc)),
                              conc=unlist(df_conc)) |> 
                              mutate(conc = ifelse(conc == "Inequitable exposure to externalities", "Ineq. externalities", conc),
                                     conc = ifelse(conc == "Sufficientarism / Capabilities approach", "Emerg. theories", conc),
                                     conc = ifelse(conc == "Transport-related exclusion", "TR exclusion", conc))

@@

NUM_standards_conc <- standards_conc |> group_by(id) |> summarize(count = n())
NUM_ENTRIES_standards_conc <- NUM_standards_conc$count |> sum()

@@

```{r, message=FALSE}
standards_conc <- standards_conc |> 
  group_by(conc, std_mcat) |> summarize(total = n())

standards_conc <- standards_conc |>
  mutate(total_perc_in_conc = total/sum(total),
         total_perc_in_all = total/NUM_ENTRIES_standards_conc)
sums_concs <- standards_conc |> group_by(conc) |> summarize(
  total_perc_in_all = sum(total_perc_in_all),
  total_perc_in_conc = 1) |> mutate(std_mcat = "TOTALS")

@@

sums_std_mcat <- standards_conc |> group_by(std_mcat) |> summarize(total_perc_in_all = sum(total_perc_in_all),
  total_perc_in_conc = 1) |> mutate(conc = "TOTALS")

@@

```{r}
set_flextable_defaults(font.size= 6,
                       font.family = NULL,
                       text.align = 'justify')

totals_conc_table_flex <- 
  # compose(totals_conc_table_flex,
  #         i = 1, j = c(1,2), part = "header", value = as_paragraph(c("Conceptualizations", "Conceptualizations"))) |>
  prepend_chunks(flextable(totals_standards_conc_table |> select(-c("Standards"))),
                 part = "body", i = c(1,2,3,4,5), j = 1, as_b(c("Vertical equity: ", "TR exclusion: ", "Wellbeing: ", "Spatial equity: ", "Horizontal equity: "))) |> 
  prepend_chunks(part = "body", i = c(1,2,3,4), j = 2, as_b(c("Ineq. externalities: ", "Rights: ", "Emerg. theories: ", "Utilitarian: "))) |> 
  # merge_h(part="header", i=1) |> #header merge
  merge_at(i=5, j=1:2) |>
  # align( align="center", part="header") |>
  delete_part("header") |>
  border_outer() |>
  border_inner_h() |>
  autofit()

totals_standards_table_flex <- 
  # compose(totals_standards_table_flex,
  #         i = 1, j = c(1), part = "header", value = as_paragraph(c("Standards"))) |> 
  prepend_chunks(flextable(totals_standards_conc_table |> select("Standards")),
                 part = "body", i = c(1,2,3,4), j = 1, as_b(c("Population: ", "Opportunity: ", "Infrastructure: ", "Environmental +: "))) |>
  merge_at(i= 4:5, j= 1) |>
  delete_part("header")|>
  border_outer() |>
  border_inner_h() |>
  autofit()
```

```{r}
standards_conc_percentages_plot <- ggplot(totals_standards_conc, 
       aes(x=std_mcat, y=conc, fill= total_perc_in_all 
           #width = rep(10, each=50)
           ))+ 
  geom_tile() + 
  scale_fill_gradient2(low = "#AFAFED", mid= "#3F3FD4", high = "#191970", midpoint = 0.2, limits = c(0,.4), na.value = "#FCB5AC") +
  scale_x_discrete(limits=(totals_standards_conc$std_mcat)[order(totals_standards_conc$group)][1:5]) + 
  scale_y_discrete(limits=(totals_standards_conc$conc)[order(totals_standards_conc$group)] ) +
  geom_text(aes(label = total_perc_in_all |> percent(accuracy = 0.1)), color = "white", size = 2.5) +
  geom_vline(xintercept=4.5, linewidth=1.5, color = "#191970") +
  geom_hline(yintercept = 43.5, linewidth=1.5, color="#191970")+
  theme_classic() +
  coord_flip()+
  theme(legend.position = "none",
        axis.ticks = element_blank(),
        axis.text.x=element_text(angle=20,hjust=1, colour = "black"),
        axis.text = element_text(size=7, colour = "black"),
        axis.title = element_blank())
#+ ggtext::geom_richtext(aes(x=5.385, y=23.5, label="Conceptualizations"), fill="grey", size=2) +
#  ggtext::geom_richtext(aes(x=3, y=49, label="Standards"), angle = 270, fill="grey", size=2) 
# standards_conc_percentages_plot
# ggsave("figures/standards_conc_percentages_plot.png")
```

@@

@@

Modern societies provide people with varying degrees of mobility and accessibility, from low wage Amazon workers’ depending on infrequent bus service, to Jeff Bezos taking private flights into space. Societies implicitly tolerate different levels of inequality in access, though these levels are occasionally made legible through policy standards. For example, U.S. transit agencies are required to set and adhere to maximum levels of disparity in service quality between minority and non-minority routes as defined by the U.S. Civil Rights Act. In contrast to prior reviews on how to measure inequality in transportation systems, this report is concerned with the implied or explicit standards that are used to judge whether the measured inequalities are fundamentally “fair” or unacceptable. The present study achieves this objective by scanning the state of academic knowledge in defining “fairness” in the transport domain.

@@

**Methods**

The authors conducted a scoping review of the literature. The review involves entering combinations of equity-related keywords into academic databases to identify relevant studies. The team initially discovered 6,832 references from three diverse databases that they shrunk to a final corpus of studies using a four-step process:

1.	A total of 611 duplicate studies were removed.  
2.	The researchers, supported by a team of trained research assistants, reviewed the remaining titles and abstracts for relevance, with 4,511 studies excluded for not sufficiently addressing transportation equity.  
3.	The full-texts of the remaining 1,711 studies were reviewed. At least two trained research assistants reviewed each document, with one of the the main researchers (listed as authors) breaking any ties. This resulted in 1,545 additional studies being excluded. At this stage, studies could be excluded if they did not include any standard for, or conceptual grounding for, equity.  
4.	The researchers examined the final corpus of 165 studies to extract data on the "When", "Where", "Who", "What", and "How" of transport equity in each study. This included annotating the methods used, standards suggested, and inferring conceptualizations of equity that may connect to broader notions of justice.

**Results**

**When and Where: the contexts of justice**. Most (60%) studies that deployed standards for equity stem from the Global North, including cases from Europe, the U.S., Canada, Australia, New Zealand, Japan, and Israel. The focus of these studies is on groups seen as disadvantaged within their respective national contexts, and strive to relate standards to national policies or issues. In contrast, research stemming from the Global South tends to adopt international standards in analyses, such as the U.N.’s Sustainable Development Goals (SDGs). Within the Global South, studies most commonly cover Asia, followed by Latin America. Studies in the latter region focus disproportionately on mobility barriers and the financial affordability of transportation. Studies pertaining to Africa are numbered, though they cover a broader scope of issues including informal transportation, citizens’ accessibility needs, and infrastructure development needs. Across the globe, the studies reviewed apply equity standards overwhelmingly to urban contexts (85%) and are concerned with a focused set of issues around the need for and the impacts of daily mobility. Studies on rural contexts are broader in the topics examined, from the equity of ferry service in the Philippines to the impacts of road construction on remote Amazonian communities.

@@

<!--We suggest: Who and What: the tools of mobility. I like our more-->
**What: the tools of mobility.** Most studies that deploy equity standards apply them to either transit (37%) or pedestrian modes (25%). Many transit analyses are multimodal in nature, comparing accessibility by transit to accessibility by automobiles as an implicit standard. Others combine transit with non-auto modes to examine access for households without vehicles. These studies focus on delivering equitable mobility, sometimes in context of also meeting climate goals. However, they often fail to include explicit minima of service that should be provided to different populations. In contrast, rights-based, infrastructure-focused studies apply principles of universal design in the study of transport systems and related infrastructure.

<!--we suggest: What: the benefits of mobility and accessibility. I like ours more; there is a distinction. -->
**What: the benefits of mobility.** Studies measure either movement (i.e., trips taken, trip quality, etc.) or the potential for movement (i.e. accessibility to specific destinations, or generally), with most studies focused on the latter. Over a quarter of the papers that focus on accessibility do not consider specific destinations, but instead concentrate on the accessibility of the trip (i.e., level of service, barriers of infrastructure) or bundles of destinations. Other studies are centred on particular destinations considered as essential, including employment, healthcare, grocery stores, and greenspace. Access to community support or places of worship, as well as childcare activities are far less common in the literature reviewed.

<!--we suggest: "How": concepts of justice and standards of fairness. I like ours more - there is a distinction. -->
**How: concepts and standards of justice.** The most common standards found in the literature, which we term opportunity standards, are defined using cut offs in access to opportunities (37.2%). Almost as common is the formulation of standards that use thresholds related to the perceptions or activities of specific populations (35.8%); we call these population standards. The latter often includes standards informed by public health, such as the number of physical activities undertaken in a week and perceived life satisfaction. Opportunity standards are evaluated through horizontal equity and spatial equity lenses, while population ones are evaluated through well-being frameworks. Another category of standards that emerged from the literature are based on infrastructure, and these include adherence to universal design guidelines or specific benchmarks for road network characteristics. Standards of thsi kind are comparatively less common in the literature and are often framed in terms of rights, particularly as they relate to achieving universal design. Infrastructure standards around sustainable infrastructure (i.e. block lengths or distance between transit stops) are also evaluated through vertical, horizontal and spatial lenses. The distribution of studies with respect to the kinds of standards deployed, and the broader concepts of justice used, are presented in the Figure below:

@@

The review conducted for this report provides a snapshot of the state of knowledge today in terms of how standards of fairness are developed and deployed. The review is telling both in terms of what it reveals about the literature, what is covered and what is not. Based on these, the authors challenge practitioners, researchers, advocates, and governments to commit to five calls to action:

@@

The standards of equity used in research are often weakly linked to any legally, intellectually, or publicly accepted concepts of justice. Many standards are arbitrary. For example, the 15-minute city sets a threshold that is easy to remember, but is this really what is needed to achieve sufficient sustainable mobility for urban residents, as opposed to a 10 or 20 minute city?  And what conceptualization of human need should determine what services are essential enough to be included within this threshold, and for whom?

@@

Mixed-methods research can help governments set meaningful standards by better linking quantitative thresholds with qualitative assessments of well-being, as well as freedom of movement and perceived access, among other facets of transportation. Yet most of the research reviewed is purely quantitative or, to a much lesser degree, purely qualitative. The constant deployment of metrics Lorenz Curves of Gini coefficients without research on where such measures need to be to improve quality of life (i.e., a standard), will not help governments with limited resources achieve the best possible equity outcomes within their respective contexts. Mobilizing Justice’s Priority Populations working group was inaugurated to advance new research in this area.

<!-- We suggested:
Call 3: Data availability as a matter of justice-->
Call 3: Making data available is a matter of justice.

Several areas of critical importance, such as access to care and community, are under-researched, and the authors suspect this relates to lack of data on these kinds of destinations. Furthermore, data issues prevent the application of many methods of equity analyses in the Global South. The production, availability, transparency, and ownership of data is a matter of justice on which governments and NGOs should act. In the Canadian context, Mobilizing Justice’s National Survey working group is working to advance data practices by conducting the first national survey on transportation poverty and striving to make data available in the form of Open Data Products.

@@

Linking standards to lived experiences and individual outcomes can help increase the likelihood of their adoption by governments. Further, they can help decision makers understand the role of transportation in improving policy outcomes in other domains like nutrition, physical and mental health, education, and human development.  These links can help make the standards used in research and policy more comprehensible and relatable to community members and people with lived experience navigating unequal mobility systems. Mobilizing Justice launched it’s Data Driven Standards working group to advance research in this area.

@@

Approximately 10% of the studies reviewed aimed to assess interventions. Delivering new equity-focused policies and programs that do not actually advance equity harms communities, metastasizes mistrust, and delegitimates future efforts to promote transportation justice. Furthermore, policies that benefit one group of people in one context may not benefit other groups in other situations. A deep breadth of equity evaluations is needed to avoid failure and guide policymakers towards interventions that will work for the communities they serve. Mobilizing Justice launched its Innovative Pilots working group to begin filling this gap, through further work is needed.

@@

This is your guide to scanning this document.

If you are primarily interested in the motivation behind this report, read the first section, [Just transportation?](#sect1)

@@

- **Accessibility**: the potential to reach destinations.

<!-- 
- **Capabilities approach**:  
-->

- **Equality**: a state where every group is treated identically.

- **Equity**: the distribution of benefits and burdens of things among the population, with a particular emphasis on those with the least advantage and quality in their outcomes.

- **Equity standards**: tools for answering the question of how to distribute the burdens and benefits and to whom? When operationalized standards effectively define what is fair.

- **Fairness**: a state of being that is achieved when a collective and individuals are treated equally and without bias as defined by their morals and values.

- **Justice**: A political ideal whereby people give and receive what they are due.

- **The temporal-spatial context of justice**: the answers to "Where?" and "When?". What regions, within what countries, spatial contexts, point of time, and processes of planning are impacting transport systems?

- **The subjects of justice**: the answer to "Who?" is the focus of justice.

- **The objects of justice**: the answer to "What?" is the focus of justice; what are the things that characterise (in)justice.

- **The methods for justice determination**: the answer to "How?" is justice defined and markers towards it tracked.

- **The rationale to justice**: the "Why?". The rationale connects with the context, subject, object and methods used for justice determination.

- **Mobility**: The potential to move.

- **Policy**: A deliberate system of principles to guide decisions and achieve desired outcomes.

<!--
- **Social exclusion**:  
-->

- **Space-time convergence**: the main product of a transportation system is to reduce the time needed to cover a unit of distance; this effect makes space and time converge.

<!--
- **Sufficientarianism**:

- **Theories of justice**:

- **Utilitarianism**:  
-->

@@

# Just transportation? {#sect1}

@@

The fiery wake of a rocket could be seen ascending, piercing the sky above West Texas. It was the morning of July 20, 2021, and on board of the rocket was a small group of four passengers that included Jeff Bezos, the founder of Amazon and then the world's richest person [@harwoodBezos2021]. The mission that day was among the first-ever private suborbital passenger flights, and the adventure (described as "intense" by one of the passengers) lasted a total of 10 minutes and 10 seconds [@harwoodBezos2021]. In addition to intense, the undertaking was expensive: a seat for the flight had previously been auctioned for no less than $28 million USD [@griffinPerson2021]. Before it was even 10:00 am (EDT), Bezos was back from his excursion, and he took time to declare that this was the "[b]est day ever". To reporters covering the event he said "I want to thank every Amazon employee and every Amazon customer, because you guys paid for all of this" [@johnsonRoadToSpace2021].

Meanwhile, firmly grounded on planet Earth, the employees whom Bezos thanked for his suborbital jaunt were struggling with some very mundane problems of their own, and none as lofty as conflicting schedules that prevented them from flying in rockets [@griffinPerson2021]. According to reports, people employed directly or indirectly by Amazon for warehouse or delivery work had, for years, been treated to "inhumane" conditions [@fungEuropean2018; @scottAmazon2019; @greeneAlgorithm2021], and subjected to surveillance on the job, degrading schedules that drove drivers to urinate in bottles, crushing demands for productivity quotas that led to injury, all the while facing little or no job security. From this perspective, they were treated less with gratitude and more as disposable inputs to feed Amazon's earnings and consumers' demands [@tungAmazon2020; @BBCNews2021; @reeseTeamsters2022; @middletonRobot2023].

Coverage of the July 20 launch by the mainstream media was in many cases uncritical. "We're going to build a road to space so that our kids and their kids can build a future" Bezos declared, before adding "...we need to do that to solve the problems here on Earth" [@johnsonRoadToSpace2021]. Few reporters saw it fit to ask what problems the billionaire planned to solve on Earth, or what kind of future Bezos was trying to build, and for whose children. In other words, the billionaire was not confronted with questions about what his trip did _for whom_ and _to whom_. A less dispassionate observer might have been excused for wondering (possibly aloud) about the basic _fairness_ of a man amassing a nigh unimaginable fortune that allowed him to build and fly his own spaceship, while masses of his employees were treated as throwaway cogs in the vast apparatus of his emporium^[Not unlike in the empire of another space billionaire, where things feel like the future except for the lives of workers [@wongTesla2017], and everything soars---including worker injuries [@taylorSpaceX2023].].
